---
title: "R Image Art"
output: 
  html_document:
      self_contained: false
      css: style.css
      includes:
        before_body: html/header.html
        after_body: html/footer.html
        in_header: html/header_content.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
library(knitr)
library(kableExtra) # tables
```

I recently showed my students an <a href="http://mfviz.com/triangulate/" target="_blank">image-art project</a> that I had worked on using D3.js, and on of them asked, 

> _Can you do that with R?_

So, I figured it was worth a try. This tutorial walks through the R code necessary to use random sampling to generate data art from image files in R. Read on, or check out the code <a href="https://github.com/mkfreeman/r-image-art" target="_blank">on GitHub</a>.

## Loading Images
In order to do image manipulation in R, you'll (of course), need to load an image into your environment. I found that the `imagr` package was well suited for this type of loading (and other types of image processing -- see the <a href="https://cran.r-project.org/web/packages/imager/vignettes/gettingstarted.html" target="_blank">documentation</a> for more information):

```{r load-image}
# Load an image into R (make sure your working directory is set!)

# Packages (for the entire project)
library(imager)    # image loading and processing
library(dplyr)     # data manipulation
library(ggplot2)   # data visualization
library(tidyr)     # data wrangling
library(ggvoronoi) # visualization

# Load an image into R
img <- load.image(file = "imgs/spain-graffiti-cropped.jpg")

# Print the image object out
print(img)
```

To view the image itself, you can use the `plot()` function (this image was downloaded from <a href="https://pixabay.com/photos/spain-vitoria-gasteiz-graffiti-88385/" target="_blank">pixabay</a>):

```{r display-image}
# View the image
plot(img)
```

In order to manipulate this image in a data-driven way, we'll need to represent it as a **data frame**. Luckily, we can do this using the `as.data.frame()` method to accomplish this:

```{r image-to-data}
# Represent the image as a data frame
img_df <- as.data.frame(img)

# Show a table of the first 10 rows of the data frame
img_df %>% 
  arrange(x, y, cc) %>% # sort by columns for viewing
  filter(row_number() < 10) %>% # Select top 10 columns
  kable("html") %>%  # Display table in R Markdown
  kable_styling(full_width = F) # Don't take up full width
```

These four columns describe how to draw the image:

- `x`: The horizontal position of a point (pixel) of the image
- `y`: The vertical position of a point (pixel) of the image
- `cc`: The color channel being represented: 1 (red), 2 (green), 3 (blue)
- `value`: The value of the color channel, on a scale from 0 to 1. 

Thus, the first three rows describe the color of the point in the top left corner of the image. Using the `rbg` function, this color is `rgb(0.788, 0.537, 0.294)`, which is <span style="background-color:`r rgb(0.788, 0.537, 0.294)`">this color<span>. 

## Image Data Manipulation
In order to draw the image more easily using familiar graphing packages (such as `ggplot2`), we'll need to reshape this data frame so that each row represents a single pixel. This can be done using the helpful `spread()` function from the `tidyr` package:

```{r comput-color}
# Add more expressive labels to the colors
img_df <- img_df %>% 
  mutate(channel = case_when(
    cc == 1 ~ "Red",
    cc == 2 ~ "Green", 
    cc == 3 ~ "Blue"
  ))

# Reshape the data frame so that each row is a point
img_wide <- img_df %>%
  select(x, y, channel, value) %>%
  spread(key = channel, value = value) %>%
  mutate(
    color = rgb(Red, Green, Blue)
  )
```

Using the `rgb()` function above, we're able to compute the `color` for each point by combining the specified amounts of `Red`, `Green`, and `Blue` (resulting in this data frame):

```{r display-table, echo = F}
img_wide %>% 
  arrange(x, y) %>% # sort by columns for viewing
  filter(row_number() < 10) %>% # Select top 10 columns
  kable("html") %>% # Display table in R Markdown
  kable_styling(full_width = F) # Don't take up full width
```

## `ggplot2` Rendering
Now that we have a single row for each point, we should be able to re-create the image using `ggplot2`:

```{r ggplot-dispaly}
# Plot points at each sampled location
ggplot(img_wide) +
  geom_point(mapping = aes(x = x, y = y, color = color)) +
  scale_color_identity() # use the actual value in the `color` column
```

Using a few plotting options, we can remove the axes and orient the image properly:

```{r plot-formatted}
ggplot(img_wide) +
  geom_point(mapping = aes(x = x, y = y, color = color)) +
  scale_color_identity() + # use the actual value in the `color` column
  scale_y_reverse() + # Orient the image properly (it's upside down!)
  theme_void() # Remove axes, background
```


## Random Sampling
Now, if we want to begin using randomness to create an abstract or artistic(?) representation of  the image, we'll have to sample only specific rows of the data:

```{r plot-sample}
# Take a sample of rows from the data frame
sample_size <- 2000
img_sample <- img_wide[sample(nrow(img_wide), sample_size), ]

# Plot only the sampled points
ggplot(img_sample) +
  geom_point(mapping = aes(x = x, y = y, color = color)) +
  scale_color_identity() + # use the actual value in the `color` column
  scale_y_reverse() + # Orient the image properly (it's upside down!)
  theme_void() # Remove axes, background
```

We can add another dimension of randomness by distorting the size using a randomly drawn value:
```{r}
# Create random weights for point size
img_sample$size <- runif(sample_size)

# Plot only the sampled points
ggplot(img_sample) +
  geom_point(mapping = aes(x = x, y = y, color = color, size = size)) +
  guides(size = FALSE) + # don't show the legend
  scale_color_identity() + # use the actual value in the `color` column
  scale_y_reverse() + # Orient the image properly (it's upside down!)
  theme_void() # Remove axes, background
```

Alternatively, you could use an element of the data (such as the amount of `Blue` in each point) to determine the size:

```{r}
# Use the amount of blue present in each point to determine the size
ggplot(img_sample) +
  geom_point(mapping = aes(x = x, y = y, color = color, size = Blue)) +
  guides(size = FALSE) + # don't show the legend
  scale_color_identity() + # use the actual value in the `color` column
  scale_y_reverse() + # Orient the image properly (it's upside down!)
  theme_void() # Remove axes, background
```

## Voronoi Diagrams
A **Voronoi Diagram** is "a partitioning of a plane into regions based on distance to points in a specific subset of the plane" (<a href="https://en.wikipedia.org/wiki/Voronoi_diagram" target="_blank">Wikipedia</a>). In other words, each area on the plane is determined by the nearest point. We can use this method (specifically, the `ggvoronoi` package) to represent the same image as a series of _areas_:

```{r}
# Create a Voronoi Diagram of the sampled points
ggplot(img_sample) +
  geom_voronoi(mapping = aes(x = x, y = y, fill = color)) +
  scale_fill_identity() +
  scale_y_reverse() +
  theme_void()

```

## Edge detection
To amplify patterns at edges of objects in the image, we can use the `cannyEdges` function to detect edges in the image:

```{r detect-edges}
# Detect edges in the image
edges <- cannyEdges(img)

# Display the edges
plot(edges)
```

We can then again leverage a data frame representation of the image (in this case, the _edges_) to sample the image more heavily in these contrast areas. In order to do so, we'll need to extract the data as a data frame and **join** it to our image data:

```{r edge-data}
# Convert the edge image to a data frame for manipulation
edges_df <- edges %>%  
  as.data.frame() %>% 
  select(x, y) %>% # only select columns of interest
  distinct(x, y) %>% # remove duplicates
  mutate(edge = 1) # indicate that these observations represent an edge
```

Once we have this representation of the data, we can use it to weight the sampling of points in the image:


```{r join-edges}
# Join on the edges data
img_wide <- img_wide %>% 
  left_join(edges_df)

# Apply a low weight to the non-edge points
img_wide$edge[is.na(img_wide$edge)] <- .05

# Re-sample from the image, applying a higher probability to the edge points
img_edge_sample <- img_wide[sample(nrow(img_wide), sample_size, prob = img_wide$edge), ]
```

Using these re-sampled edges, we can create more intensified images:

```{r final-plots}
# Re-create the voronoi diagram with the re-sampled data
ggplot(img_edge_sample) +
  geom_voronoi(mapping = aes(x = x, y = y, fill = color)) +
  scale_fill_identity() +
  guides(fill = FALSE) +
  scale_y_reverse() +
  theme_void() # Remove axes, background

# Re-create the scatter plot with the re-sampled data (add random sizing of circles)
ggplot(img_edge_sample) +
  geom_point(mapping = aes(x = x, y = y, color = color, size = edge * runif(sample_size))) +
  guides(fill = FALSE, size= FALSE) +
  scale_color_identity() +
  scale_y_reverse() +
  theme_void() # Remove axes, background

```

## About
This explanation was built by <a href="http://mfviz.com/" target="_blank">Michael Freeman</a>, a faculty member at the University of Washington <a href="https://ischool.uw.edu/" target="_blank">Information School</a>.

If you like this explanation (and are excited about learning `R` code) check out my new book, <a href="https://www.amazon.com/gp/product/0135133106" target="_blank">Programming Skills for Data Science</a>.
